\Huge Stellar continuum modelling

### 2023-12-18 paper/main.tex
update abstract with variants

\begin{abstract}\noindent
Continuum normalization is often a necessary step when analyzing stellar spectra. Best practice is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but computing stellar spectra can be expensive.
We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption in theoretical (normalized) spectra, and a sines-and-cosines basis for the joint continuum-instrument response.
The non-negative matrix factorization ensures that eigenspectra are strictly additive, thereby restricting the predicted normalized stellar flux to be at most 1, with the joint continuum-instrument response component to describe the rest.
The linearity of both components ensures that inference is stable and fast.
We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of all evolutionary stages: from M-dwarfs to white dwarfs.
We find good fits to observed spectra, even in situations where the theoretical grid used to approximate line absorption is provably incorrect for some features.
%The stellar absorption model is built from a non-negative matrix factorization of a theoretical continuum-normalized spectra, which ensures it cannot predict spectra more than 1.
%The constraints restrict the stellar absorption to 
We show that we obtain unbiased estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
This approach largely eliminates subjectivity in stellar continuum normalization.
\end{abstract}

@@

Continuum normalisation is a necessary step

@@

% NMF first
The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a grid of $N$ theoretical spectra each with $D$  continuum-normalized fluxes, which we denote as $\vec{M}$. In the applications that we present, the data are assumed to have the same wavelength sampling and a consistent line spread function, such that we use a grid where $D = P$. Other situations are permitted, but at inference time there is a need to interpolate the line absorption to the $P$ observed pixels. With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecX$ to be
\begin{align}
    \vecX = 1 - \vec{M}
\end{align}
such that $\vecX \in \left[0, 1\right)$ and is zero when no line absorption exists. This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecX$ using non-negative matrix factorization (NMF) such that 
\begin{align}
    \vecX \approx \vec{W}\vec{H} \label{eq:nmf}
\end{align}
where all elements in $\vec{W}$ and $\vec{H}$ are non-negative. 
We could construct an approximation directly to $\vec{M}$, but this would lead to much denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX$ that are 0 (i.e., no line absorption). For this approximation we must select the number of components $C$ to use, which is significantly smaller than both the number of theoretical spectra $N$ and the number of pixels $D$ per theoretical spectrum. Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding eigenspectra. Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$. \\

@@

@@

\begin{figure*}
    \caption{The median pixel $\chi^2$ value as a function of \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence stars observed with the BOSS spectrograph. \todo{Should expect to see increasing residuals due to TiO bands not captured by the model, and emission lines.}}
\end{figure*}

@@

@@

\paragraph{Acknowledgements}
It is a pleasure to thank
% All these people are possible co-authors
    Adam Wheeler (Ohio State University),
    David W. Hogg (New York University),
    Megan Bedell (Flatiron Institute),
.
% include bibliography
\bibliographystyle{aasjournal}
%\bibliography{bibliography}

>>>>

\begin{abstract}\noindent
Continuum normalization is often a necessary step when analyzing stellar spectra. 
%The best approach is to forward model the continuum (and instrument response) simultaneously with the stellar parameters, but continuum normalization is usually required before estimating the stellar parameters.
We present a forward model to simultaneously fit stellar absorption and the joint continuum-instrument response using two linear models: a non-negative matrix factorization to approximate line absorption,
% in theoretical (normalized) spectra, 
and a sines-and-cosines basis for the joint continuum-instrument response.
The non-negative matrix factorization ensures that basis spectra are strictly additive, thereby restricting the predicted normalized flux to be at most 1, with the sines-and-cosines basis to describe the rest.
The linearity of both components ensures that inference is stable and fast.
We describe three variants of this approach:
    (1) a model-driven approach where line absorption is factorized entirely from continuum-normalized theoretical spectra;
    (2) a purely data-driven approach where line absorption and continuum are fit entirely from data;
    and
    (3) a hybrid approach where basis spectra are initialized from factorization of theoretical spectra, but iteratively adjusted to fit data. 
We apply these methods to \project{Sloan Digital Sky Survey} optical and infrared spectra of stars spanning evolutionary stages from pre-main-sequence stars to white dwarfs.
We find good fits to data, even for the model-driven variant where theoretical spectra used to factorize line absorption is provably incorrect for some features.
We show that the basis amplitudes can be used to reliably estimate stellar parameters and chemical abundances, 
and that we obtain low bias estimates of the continuum as a function of stellar parameters and signal-to-noise ratios.
%This approach largely eliminates subjectivity in stellar continuum normalization.
\end{abstract}

@@

Continuum normalization is a subject
Continuum normalisation is a necessary step

@@

% NMF first
The line absorption model $f(\lambda_i;\vecpsi)$ is constructed from a grid of $N$ theoretical spectra each with $D$  continuum-normalized fluxes, which we denote as $\vec{M}$. In the applications that we present, the data are assumed to have the same wavelength sampling and a consistent line spread function, such that we use a grid where $D = P$. Other situations are permitted, but at inference time there is a need to interpolate the line absorption to the $P$ observed pixels. With our $\vec{M}$ matrix of continuum-normalized theoretical flux values, we then defined \emph{stellar absorption} $\vecX$ to be
\begin{align}
    \vecX = 1 - \vec{M}
\end{align}
such that $\vecX \in \left[0, 1\right)$ and is zero when no line absorption exists. This is a `trick' that allows us to construct a highly constrained and sparse approximation to the matrix $\vecX$ using non-negative matrix factorization (NMF) such that 
\begin{align}
    \vecX \approx \vec{W}\vec{H} \label{eq:nmf}
\end{align}
where all elements in $\vecX$, $\vec{W}$, and $\vec{H}$ are required to be non-negative. 
We could construct an approximation directly to $\vec{M}$, but this would lead to denser matrices: there are \emph{no} elements of $\vec{M}$ that are exactly 0 (i.e., total line absorption), but there are \emph{many} elements in $\vecX$ that are 0 (i.e., no line absorption). For this approximation we must select the number of components $C$ to use, which is significantly smaller than both the number of theoretical spectra $N$ and the number of pixels $D$ per theoretical spectrum. Here $\vec{W}$ is a $N \times C$ matrix that can be thought of as $C$ amplitudes per theoretical spectrum, and $\vec{H}$ is a $C \times D$ matrix of $C$ corresponding eigenspectra. Figure~\ref{fig:schematic} illustrates the NMF procedure and shows some example infrared eigenspectra $\vec{H}$. \\

@@

\begin{figure}
    \caption{A Gaia H-R diagram showing the median $\chi^2$ per bin.}
\end{figure}

@@

\begin{figure*}
    \caption{The median pixel $\chi^2$ value as a function of \emph{Gaia} $\mathrm{BP} - \mathrm{RP}$ color for main-sequence stars observed with the BOSS spectrograph. \todo{Should expect to see increasing residuals due to TiO bands not captured by the model, and emission lines.}}
    % Could do the same showing standard deviation per bin
\end{figure*}

@@

\section{Discussion}\label{sec:discussion}

- getting good fits even with bad models: the hydrogen lines in the apogee grid are knowingly incorrect, but we see no difference in the $\chi^2$ at those wavelengths. That is because the hydrogen lines separate out into very clean eigenspectra, where the amplitude for that eigenspectra is not required to vary smoothly with other amplitudes

- biases as a function of S/N or labels

- poorly modelled regions as a function of pixel

- $\chi^2$ across the HRD

- excellent way to identify regions where models and observations may disagree

- orthogonality of the two model components

- issue with M-giant eigenspectra when they were over-represented in the sample

@@

\paragraph{Acknowledgements}
It is a pleasure to thank
% All these people are possible co-authors
    Adam Wheeler (Ohio State University),
    Andrew Saydjari (Harvard),
    David W. Hogg (New York University),
    Megan Bedell (Flatiron Institute),
.
% include bibliography
\bibliographystyle{aasjournal}
%\bibliography{bibliography}
